---
title: "Custom instruments in tallieR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Custom instruments in tallieR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

tallieR includes 27 built-in instruments, but many studies use questionnaires that are not in the package --- proprietary tools, study-specific scales, or newer measures that haven't been reviewed for inclusion yet. The custom instrument system lets you load any questionnaire defined as a ScoreMe-compatible JSON spec and use it through the same API as built-in instruments, without modifying tallieR source code.

## The JSON spec format

A custom instrument is defined in a single `.json` file. The required fields are `id`, `title`, and `items`. Everything else is optional but recommended.

tallieR ships with a worked example --- the Fatigue Severity Scale (FSS):

```{r fss-path}
library(tallieR)

fss_path <- system.file("extdata", "example_instrument.json", package = "tallieR")
```

The spec structure looks like this (abbreviated):

```json
{
  "id": "fss",
  "title": "Fatigue Severity Scale",
  "domain": "Fatigue",
  "maxScore": 63,
  "scoringMethod": {
    "type": "mean",
    "items": ["fss1", "fss2", "fss3", "fss4", "fss5", "fss6", "fss7", "fss8", "fss9"]
  },
  "scoreBands": [
    { "min": 1.0, "max": 3.9, "label": "No significant fatigue",
      "color": "#2E7D32", "description": "Below clinical threshold." },
    { "min": 4.0, "max": 5.4, "label": "Moderate fatigue",
      "color": "#F59E0B", "description": "May be affecting daily activities." },
    { "min": 5.5, "max": 7.0, "label": "Severe fatigue",
      "color": "#DC2626", "description": "Significant impact on functioning." }
  ],
  "items": [
    { "id": "fss1", "number": 1,
      "text": "My motivation is lower when I am fatigued.",
      "type": "scale_1_10" }
  ]
}
```

### Supported scoring types

| `scoringMethod.type` | Behaviour |
|---|---|
| `"sum"` | Sum of item values |
| `"weighted_sum"` | Sum of item values x `multiplier` |
| `"mean"` | Mean of item values (1 d.p.) x `multiplier` |
| `"composite"` | Cannot be auto-compiled --- see [Composite scoring] below |

### Key fields reference

| Field | Required | Description |
|---|---|---|
| `id` | yes | Short identifier used in tallieR (e.g. `"fss"`) |
| `title` | yes | Full instrument name |
| `items` | yes | Array of item objects, each with `id` and `type` |
| `domain` | | Clinical domain (defaults to `"Custom"`) |
| `maxScore` | | Maximum possible score |
| `beta` | | Mark as unvalidated (`true`/`false`) |
| `scoringMethod` | | Scoring algorithm definition |
| `scoreBands` | | Clinical interpretation bands (`min`, `max`, `label`, `color`, `description`) |
| `scoreBandDirection` | | `"asc"` (default) or `"desc"` |

## Loading a custom instrument

`load_instrument()` reads the JSON file, compiles scoring and interpretation functions, and returns a named registry entry:

```{r load}
fss <- load_instrument(fss_path)
```

The registry is a named list whose name matches the instrument `id`:

```{r inspect}
names(fss)
fss$fss$title
fss$fss$domain
fss$fss$max_score
```

## Scoring with a custom instrument

Pass the registry to `score_questionnaire()` via the `instruments` argument:

```{r score}
fss_answers <- list(
  fss1 = 5, fss2 = 6, fss3 = 5, fss4 = 6,
  fss5 = 5, fss6 = 4, fss7 = 5, fss8 = 4, fss9 = 5
)

score_questionnaire("fss", fss_answers, instruments = fss)
interpret_score("fss", 5.0, instruments = fss)
```

The `instruments` argument is available on all scoring and tidy functions:
`read_scoreme()`, `read_scoreme_dir()`, `score_questionnaire()`, `interpret_score()`,
`interpret_all()`, `cronbach_alpha()`, and `omega_reliability()`.

## Using custom instruments with imports

Pass the registry to `read_scoreme()` so that custom instruments in the export are scored on import:

```r
fss  <- load_instrument("instruments/fss.json")
exp  <- read_scoreme("my_export.json", instruments = fss)
wide <- scores_wide(exp)
```

## Loading multiple instruments

Combine registries with `c()`:

```r
registry <- c(
  load_instrument("instruments/fss.json"),
  load_instrument("instruments/vas_pain.json")
)
```

Or load everything in a folder at once:

```r
registry <- load_instrument_dir("instruments/")
exp      <- read_scoreme("export.json", instruments = registry)
```

## Composite scoring

Instruments with `"composite"` scoring (multi-component algorithms like PSQI) cannot be auto-compiled from the JSON spec. Loading such a spec succeeds, but scoring returns `NA` with a warning. Override the compiled `score` function after loading:

```r
my_scale <- load_instrument("composite_scale.json")

my_scale$composite_scale$score <- function(answers) {
  component_a <- as.numeric(answers[["item1"]]) + as.numeric(answers[["item2"]])
  component_b <- as.numeric(answers[["item3"]])
  component_a + component_b
}

score_questionnaire("composite_scale", answers, instruments = my_scale)
```

## Inspecting a loaded registry

Custom instruments do not appear in `available_instruments()` --- that lists only built-ins. To see what is in a custom registry:

```{r registry-inspect}
data.frame(
  id        = names(fss),
  title     = vapply(fss, `[[`, character(1), "title"),
  domain    = vapply(fss, `[[`, character(1), "domain"),
  max_score = vapply(fss, `[[`, numeric(1),   "max_score")
)
```

Custom entries take precedence over built-ins when the same `id` is used, so a custom entry can also override a built-in scoring algorithm if needed.

