---
title: "Getting started with tallieR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with tallieR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

tallieR is the R companion to the [ScoreMe app](https://scoreme.circadia-lab.uk). It imports
participant JSON exports and returns tidy data frames ready for analysis.

## Installation

```r
# install.packages("pak")
pak::pkg_install("circadia-bio/tallieR")
```

## Basic workflow

### 1. Read an export

ScoreMe can export all participants in one JSON file. Point `read_scoreme()` at it:

```{r read}
library(tallieR)

path <- system.file("extdata", "example_export.json", package = "tallieR")
exp  <- read_scoreme(path)
exp
```

Or read a whole folder of exports at once:

```r
study <- read_scoreme_dir("~/Downloads/my_study_exports/")
```

### 2. Wide score table

One row per participant, one column per questionnaire (most recent session):

```{r wide}
wide <- scores_wide(exp)
wide[, c("code", "age", "sex", "ess", "isi")]
```

### 3. Long score table

One row per participant × questionnaire × administration (full history):

```{r long}
long <- scores_long(exp)
head(long[, c("code", "questionnaire_id", "completed_at", "score")])
```

### 4. Item-level data

One row per item response — useful for reliability analysis or IRT:

```{r items}
items <- items_long(exp)
head(items[, c("code", "questionnaire_id", "item_id", "response")])
```

## Scoring and interpretation

tallieR rescores all questionnaires from raw item responses by default.
You can also call the scoring functions directly:

```{r scoring}
score_questionnaire("ess", list(
  ess1 = 2, ess2 = 1, ess3 = 0, ess4 = 3,
  ess5 = 1, ess6 = 0, ess7 = 2, ess8 = 1
))

interpret_score("ess", 10)
```

PSQI returns a named list of component scores alongside the global score:

```{r psqi}
psqi_answers <- list(
  psqi9 = 2, psqi2 = 35, psqi5a = 2, psqi4 = 5.5,
  psqi1 = list(hour = 0, minute = 30),
  psqi3 = list(hour = 7, minute = 0),
  psqi5b=2, psqi5c=1, psqi5d=0, psqi5e=1,
  psqi5f=0, psqi5g=1, psqi5h=1, psqi5i=0,
  psqi6 = 1, psqi7 = 2, psqi8 = 2
)
score_questionnaire("psqi", psqi_answers)
```

## Available instruments

```{r instruments}
available_instruments()
```

The `has_reverse` column flags instruments with item-level reverse scoring. Pass `scored_items = TRUE` to `items_long()` to apply reversals automatically.

## Study monitoring

`summary()` prints a structured overview of participant count, instruments, completion rates, and date range:

```{r summary}
summary(exp)
```

Access the stats programmatically via the invisible return value:

```{r summary-prog}
s <- summary(exp)
s$completion
```

`completion_summary()` returns a tidy data frame of completion status per participant and questionnaire:

```{r completion}
# Long format: one row per participant x questionnaire
completion_summary(exp)

# Wide format: one logical column per questionnaire
completion_summary(exp, wide = TRUE)
```

## Clinical interpretations

`interpret_all()` returns clinical interpretations for all results in one call, as a long data frame joinable with `scores_long()`:

```{r interpret-all}
interps <- interpret_all(exp)
interps[, c("code", "questionnaire_id", "score", "label")]
```

## Reverse-scored items

For instruments with reverse-scored items (currently STAI-S and STAI-T), pass `scored_items = TRUE` to get a `response_scored` column alongside the raw `response`:

```r
items_long(exp, scored_items = TRUE)
```

## Reliability analysis

`cronbach_alpha()` computes Cronbach's α with exact CIs. `omega_reliability()` computes McDonald's ω via single-factor EFA — generally preferred for non-tau-equivalent items:

```r
cronbach_alpha(exp)
omega_reliability(exp)
```

Both accept a `tallier_export`/`tallier_study` object or an `items_long()` data frame directly.

## Tidyverse integration

Any study object coerces directly to a tibble via `scores_wide()`:

```r
library(tibble)
tibble::as_tibble(exp)
```
